Place of death in NRS COVID-19 deaths: data processing for analysis, 2022 data

### 2022-02-04 covid_nrs_place_of_death_data_2022.Rmd
added hsitorical figure calculations. phew!

files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
```

@@

```{r}
ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")

weekly_deaths_overall <-
  # ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  ods_raw %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2022)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
```

@@

# TODO: complete cleaning of this bit
# TODO: canonical causes of death

# weekly_deaths_cause_of_death <-
weekly_cause_of_death_data <- 
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

```{r}

```

>>>>

files_weekly_deaths <- set_names(x = as.list(table_of_data_sources$path_latest), nm = table_of_data_sources$short_name)
files_historical_deaths <- set_names(x = as.list(table_of_historical_data_sources$path_latest), nm = table_of_historical_data_sources$short_name)
files_case_counts <- set_names(x = as.list(table_of_case_trends_data_sources$path_latest), nm = table_of_case_trends_data_sources$short_name)
```

@@

```{r}
## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")

weekly_deaths_overall <-
  ods_raw %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2022)) %>% # remove totals for 2020, 2021, 2022, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(202[0-3])")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(202[0-3])\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # not needed here probably, but no harm in including it
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing
```

@@

weekly_cause_of_death_2021_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

```{r}
historical_average_cause_of_death_2021_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
    date_w_c = as.Date(date_w_c),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
    historic_period = historic_period_depending_on_year(year)
  ) %>%
  select(-name) %>%
  drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future
```

### Data from 2020 & 2021

The reports available from NRS in 2022 include data from 2021-2022, but we can also include data from 2020 for context. For this I'll load the last 'main report' file from 2021.

The weekly causes of death data for 2020 is included in the sheet `Table 3 (2020)`.

#### 'Main report' file

```{r}
nrs_weekly_data_sheet_2020 <- 
  excel_sheets(files_historical_deaths$`2020-2021`)[which(str_detect(tolower(excel_sheets(files_historical_deaths$`2020-2021`)), pattern = "table 3.*2020"))]

## construct column names for the data
headers_in_nrs_data_2020 <-
  c(
    "place_of_death",
    "cause_of_death",
    paste0("week_", 1:53),
    "blank_remove",  # this and the rest are irrelevant entries that we'll remove
    "remove_1",
    "remove_2"
  )

## extract data... this is dense but I've tried to annotate it clearly!
weekly_deaths_cause_of_death_2020 <-
  readxl::read_excel(files_historical_deaths$`2020-2021`, sheet = nrs_weekly_data_sheet_2020, col_names = headers_in_nrs_data_2020, range = cell_rows(6:125)) %>%  # the data are inrows 6:125
  select(-matches("remove")) %>%  # remove irrelevant summary columns at the end
  mutate(across(.cols = matches("week"), .fns = as.numeric)) %>%  # remove erroneous "count1, count2, ..." entries due to superscript in previous column
  mutate(dataset = if_else(condition = str_detect(string = cause_of_death, pattern = regex(pattern = "^registered|^difference", ignore_case = TRUE)), true = cause_of_death, false = NA_character_)) %>%  # extract the Registered deaths: historical, 2020/2021, and Difference sections
  tidyr::fill(place_of_death, .direction = "down") %>%  # complete the entries for place of death from the "last observed"
  tidyr::fill(dataset, .direction = "down") %>%  # same as above but for "dataset" column
  drop_na(week_1) %>%  # remove blank "header" rows with no data recorded
  pivot_longer(cols = matches("week"), names_to = "week_number", values_to = "number_of_deaths", values_drop_na = TRUE) %>%  # convert to long data format
  mutate(
    week_number = as.integer(parse_number(week_number)),  # extract week number
    place_of_death =  # find best match in the reference names used in other datasets
      order_of_place_of_death_levels[stringdist::amatch(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels,1,4), method = "lv", maxDist = 5)],  # the reference is in the variable order_of_place_of_death_levels
    year = 2020,
    dataset = str_remove_all(dataset , "Registered deaths\\: "),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
    ) %>%
  filter(dataset!="Difference") %>%
  pivot_wider(names_from = dataset, values_from = number_of_deaths) %>%
  rename(deaths_mean = `five year average`, number_of_deaths = `2020`)
```

#### Merge 2020 data with current data

Note: the 5-year average for 2022 (spanning 2016-2019 & 2021) is only published along with the running weekly counts of deaths by cause of death; it's just that NRS has decided to only publish them at the same time as new data comes out! This is a bit of a bummer - it means we can't populate the whole year of 'historical context' data for 2022.

They could be calculated by hand if the 2015-2019 weekly data were available broken down by cause of death - as it stands, only the average 2015-2019 figure is available.

```{r}
weekly_cause_of_death_data <-
  bind_rows(
    weekly_cause_of_death_2021_onwards,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
  )

historical_average_cause_of_death <-
  bind_rows(
    historical_average_cause_of_death_2021_onwards,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-number_of_deaths) %>% 
      mutate(historic_period = historic_period_depending_on_year(year))
  )
```

## Historical weekly deaths

The format of the .xlsx files is that sheet 1 is a table of contents, sheet 2 contains the data, and sheet 3 is a pivot table for subsetting data. These are downloaded from the ad-hoc section also.

```{r, warning=FALSE}
weekly_deaths_2015_2019_overall <- 
  read_csv(files_historical_deaths$overall, skip = 2) %>%
  rename(year = 1) %>%  # rename first column
  mutate(
    place_of_death = if_else(is.na(`1`), true = year, NA_character_)
  ) %>%
  fill(place_of_death, .direction="down") %>%  # LOCF on place_of_death
  drop_na(`1`) %>%
  pivot_longer(cols = all_of(as.character(1:53)), values_to = "number_of_deaths", names_to = "week_number") %>%
  mutate(week_number=as.integer(week_number)) %>%
  recode_place_of_death()

weekly_deaths_2015_2019_hb <- 
  readxl::read_excel(files_historical_deaths$hb, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(  # same format as current data
    week_number = matches("week"),
    place_of_death = matches("location"),
    ref_area = health_board
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

weekly_deaths_2015_2019_la <-
  readxl::read_excel(files_historical_deaths$la, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(  # same format as current data
    week_number = matches("week"),
    place_of_death = matches("location"),
    ref_area = council_area
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

weekly_deaths_2015_2019_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_historical_deaths$sex_age, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  mutate(
    year = 2000L + as.integer(year),
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()

## compute weekly data for sex and age separately
## sum over sex or age, respectively
weekly_deaths_2015_2019_sex <-
  weekly_deaths_2015_2019_sex_age %>%
  group_by_all %>%
  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")

weekly_deaths_2015_2019_age <-
  weekly_deaths_2015_2019_sex_age %>%
  group_by_all %>%
  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
```

## Map data

I'm using `sf` to read these and to include them invisualisations.

* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-december-2020-uk-buc/
* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
  - Note: this is a fairly high resolution, don't need this much detail!
  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)

```{r}
shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
shapefile_la <- st_read(dsn = "./map_data/la")
```

## COVID case counts by LA/HB

```{r}
cases_hb <- read_csv(
  files_case_counts$hb, 
  col_types = 
    cols(
    .default = col_double(),
    `_id` = col_skip(),
    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
    Date = col_date(format = "%Y%m%d"),
    HB = col_character(),
    HBName = col_character(),
    HospitalAdmissionsQF = col_character(),
    ICUAdmissionsQF = col_character()
    ),
  ) %>% clean_names() %>%
  rename(ref_area = hb_name) %>%
  mutate(
    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
    ref_area = phsmethods::match_area(hb)  # convert to standard HB names
    )

cases_la <- read_csv(
  file = files_case_counts$la, 
  col_types = 
    cols(
    .default = col_double(),
    `_id` = col_skip(),
    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
    Date = col_date(format = "%Y%m%d"),
    CA = col_character(),
    CAName = col_character()
  )
) %>% clean_names() %>%
  rename(ref_area = ca_name) %>%
  mutate(
    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8)),
    ref_area = phsmethods::match_area(ca)  # convert to standard LA names, jsut in case they weren't already!
    )
```

# Calculate case counts: weekly counts to match weekly death reporting

Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks. This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!

However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!

Note also that the `crude_rate_positive_per100k` represents the cumulative rate!

```{r}
weekly_covid_cases_hb <-
  cases_hb %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    crude_rate_positive_per100k = max(crude_rate_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )

## extract Scotland-wide data and remove from HB split
weekly_covid_cases_overall <- weekly_covid_cases_hb %>% filter(ref_area=="Scotland")
weekly_covid_cases_hb <- weekly_covid_cases_hb %>% filter(ref_area!="Scotland")

weekly_covid_cases_la <-
  cases_la %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    crude_rate_positive_per100k = max(crude_rate_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )
```

# Historical averages, minima & maxima

Due to the counting of weeks, most years have 52 weeks with some years (2015 & 2020) having 53. Therefore we need to decide how to deal with week 53 in 2020 since only 2015 has a corresponding weekly count. The solution used here is to calculate a weekly rate for week 53 from week 53 in 2015, and week 52 in 2016-2019 - this represents the "last week of the year", which includes December 31st.

Note also that since 2022, the 5-year historical period has been defined as 2016-2021, with 2020 excluded, so we need to calculate two weekly averages!

```{r, warning=FALSE}
historical_weekly_average_overall <-
  bind_rows(
    ## 2015-2019 period
    weekly_deaths_2015_2019_overall %>%
    filter(year!="average") %>%  # remove averages
    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
    filter(!all(number_of_deaths==0 & week_number==53)) %>%
    ungroup %>%
    create_week_53_data_for_years_with_52_weeks %>%
    group_by(week_number,place_of_death) %>% 
    summarise(
      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
      .groups = "drop"
    ) %>% 
    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2021, excluding 2020
    bind_rows(
      ## 2016-2019
      weekly_deaths_2015_2019_overall %>%
        mutate(year = as.integer(year)) %>%
        filter(year %in% 2016:2019),
      ## 2021
      weekly_deaths_overall %>% 
        rename(number_of_deaths = deaths_all_causes) %>%
        filter(year == 2021 & age == "all" & sex == "all")
      ) %>%
    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
    filter(!all(number_of_deaths==0 & week_number==53)) %>%
    ungroup %>%
    create_week_53_data_for_years_with_52_weeks %>%
    group_by(week_number,place_of_death) %>% 
    summarise(
      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
      .groups = "drop"
    ) %>% 
    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."

# TODO: weekly averages may be too detailed for this level od subcategorising!

deaths_mean_min_max <- function(data_tbl) {
  data_tbl %>%
    summarise(
      deaths_mean = mean(number_of_deaths),
      deaths_min = min(number_of_deaths),
      deaths_max = max(number_of_deaths),
      .groups = "drop"
    )
}

make_2016_to_2019_and_2021_data <- function(historical_data, current_data) {
  bind_rows(
      historical_data %>% filter(year %in% 2016:2019),
      current_data %>% filter(year == 2021) %>% rename(number_of_deaths = deaths_all_causes)
    )
}

historical_weekly_average_hb <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_hb %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_hb, 
      current_data = weekly_deaths_hb
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_la <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_la %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_la, 
      current_data = weekly_deaths_la
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_sex_age <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_sex_age %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_sex_age, 
      current_data = weekly_deaths_sex_age
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_age <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_age %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_age, 
      current_data = weekly_deaths_age
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_sex <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_sex %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_sex, 
      current_data = weekly_deaths_sex
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

# TODO: include the split between historical periods here also
compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}

historical_annual_average_overall <-
  bind_rows(
      ## 2016-2019
      weekly_deaths_2015_2019_overall %>%
        mutate(year = as.integer(year)) %>%
        filter(year %in% 2016:2019) %>%
        mutate(historic_period = historic_period_depending_on_year(2020)),
      ## 2021
      weekly_deaths_overall %>% 
        rename(number_of_deaths = deaths_all_causes) %>%
        filter(year == 2021 & age == "all" & sex == "all") %>%
        mutate(historic_period = historic_period_depending_on_year(2022))
      ) %>%
  compute_historical_annual_average_deaths()

make_2016_to_2019_and_2021_data_with_historic_periods <- function (historical_data, current_data) {
  bind_rows(
    historical_data %>% 
      filter(year %in% 2016:2019) %>%
      mutate(historic_period = historic_period_depending_on_year(2020)),
    current_data %>% 
      filter(year == 2021) %>% 
      rename(number_of_deaths = deaths_all_causes) %>%
      mutate(historic_period = historic_period_depending_on_year(2022))
  )
}

historical_annual_average_hb <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_hb, 
    current_data = weekly_deaths_hb
      ) %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_la <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_la, 
    current_data = weekly_deaths_la
      ) %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_sex_age <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_sex_age, 
    current_data = weekly_deaths_sex_age
      ) %>%
  compute_historical_annual_average_deaths(sex, age)

historical_annual_average_sex <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_sex, 
    current_data = weekly_deaths_sex
      ) %>%
  compute_historical_annual_average_deaths(sex)

historical_annual_average_age <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_age, 
    current_data = weekly_deaths_age
      ) %>%
  compute_historical_annual_average_deaths(age)
```

## Historical annual average by cause

I can approximate this by taking the sum of the average weekly deaths for `2015-2019`; for the `2016-2019 + 2021` period this isn't readily available because NRS only publish the weekly averages alongside the weekly updated data, but not for the entire year. Therefore I'll just calculate the `2015-2019` period average for now, and a compromise `2015-2019 + 2021` average:

$$\frac{5}{6}average\_deaths_{2015-2019} + \frac{1}{6}deaths_{2021}$$

```{r}
# TODO: incorporate the 2016-2019 + 2021 period
historical_annual_average_cause_of_death_2015_2019 <-
  historical_average_cause_of_death %>%
  filter(historic_period == "2015-2019" & year == 2021) %>%  # this is slightly strange but some of the averages disagree, probably because the 2015-2019 data were updated in 2020, and the 2021 data were published with modified figures. NRS deaths data are truly a living document!
  group_by(place_of_death, cause_of_death, historic_period) %>%
  summarise(
    annual_deaths_mean = sum(deaths_mean),
    .groups = "drop"
  )

deaths_by_cause_2021 <-
  weekly_cause_of_death_2021_onwards %>% 
  filter(year == 2021) %>% 
  group_by(place_of_death, cause_of_death) %>% 
  summarise(deaths_2021 = sum(number_of_deaths), .groups = "drop")

historical_annual_average_cause_of_death <-
  bind_rows(
    historical_annual_average_cause_of_death_2015_2019
    ,
    ## compromise figures for 2015-2019 + 2021
    historical_annual_average_cause_of_death_2015_2019 %>%
    filter(cause_of_death != "All") %>%
    left_join(
      deaths_by_cause_2021, by = c("place_of_death", "cause_of_death")
    ) %>%
    mutate(
      annual_deaths_mean = (5/6 * annual_deaths_mean) + (1/6 * deaths_2021),
      deaths_2021 = NULL,
      historic_period = "2015-2019, 2021"
    )
  )
```

## Check weekly historic averages against NRS computed data

```{r}
weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  mutate(historic_period = historic_period_depending_on_year(year)) %>%
  left_join(historical_weekly_average_overall) %>%
  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean, historic_period) %>%
  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e. all other entries had the same figures!). Week 53 was computed as the average of deaths in week 53 in 2015 and weeks 52 in 2016-2019. For all the other datasets the mean deaths computed agree!")
```
